---
title: "Cherry Blossom Peak Bloom Predictions"
author: Sara Alhassani and D.J. Grant
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Necessary Packages
```{r}
library(readr)
library(tidyr)
library(dplyr)
library(rnoaa)
library(lubridate)
library(ggplot2)
library(stringr)
```

## Importing Past Year Data

DC bloom data and restrict DC bloom data to necessary columns and years
```{r}
washingtondc <- read_csv("data/washingtondc.csv")

DC_bloom <- washingtondc %>%
  select(location, year, bloom_date, bloom_doy) %>%
  filter(year >= 2014)
```

Kyoto bloom data and restrict Kyoto bloom data to necessary columns and years
```{r}
kyoto <- read_csv("data/kyoto.csv")

Ky_bloom <- kyoto %>%
  select(location, year, bloom_date, bloom_doy) %>%
  filter(year >= 2014)
```

Liestal bloom data and restrict Liestal bloom data to necessary columns and years
```{r}
liestal <- read_csv("data/liestal.csv")

Li_bloom <- liestal %>%
  select(location, year, bloom_date, bloom_doy) %>%
  filter(year >= 2014)
```

## Importing PM 2.5 Data

DC: PM 2.5 data and Restrict variables and years
```{r}
DC_pm2.5 <- read.csv("pm data/DCpm.txt",header= TRUE, sep="", row.names=NULL)

DC_pm2.5 <- DC_pm2.5 %>%
  select("Year", "Month", "Day", "UTC_Hour", "PM2.5") %>%
  filter(Year >= 2014)

#Take daily average from the hourly 

DC_avgpm <- DC_pm2.5 %>%
  group_by(Year, Month, Day) %>%
  rename(year = Year, month = Month, day = Day) %>%
  summarize(PM_mean = mean(PM2.5))

DC_avgpm$city <- "washingtondc"
```


Kyoto: PM 2.5 data and Restrict variables and years
```{r}
Ky_pm2.5 <- read.csv("pm data/Kyotopm.txt" ,header= TRUE, sep="", row.names=NULL)

Ky_pm2.5 <- Ky_pm2.5 %>%
  select("Year", "Month", "Day", "UTC_Hour", "PM2.5") %>%
  filter(Year >= 2014)

#Take daily average from the hourly 

Ky_avgpm <- Ky_pm2.5 %>%
  group_by(Year, Month, Day)%>%
  rename(year = Year, month = Month, day = Day) %>%
  summarize(PM_mean = mean(PM2.5))

Ky_avgpm$city <- "kyoto"
```


Liestal: PM 2.5 data and Restrict variables and years
```{r}
Li_pm2.5 <- read.csv("pm data/Switzerlandpm.txt",header= TRUE, sep="", row.names=NULL)

Li_pm2.5 <- Li_pm2.5 %>%
  select("Year", "Month", "Day", "UTC_Hour", "PM2.5") %>%
  filter(Year >= 2014)

#Take daily average from the hourly 

Li_avgpm <- Li_pm2.5 %>%
  group_by(Year, Month, Day) %>%
  rename(year = Year, month = Month, day = Day) %>%
  summarize(PM_mean = mean(PM2.5))

Li_avgpm$city <- "liestal"
```


Vancouver: PM 2.5 data and Restrict variables and years
```{r}
Vc_pm2.5 <- read.csv("pm data/Vancouverpm.txt",header= TRUE, sep="", row.names=NULL)

Vc_pm2.5 <- Vc_pm2.5 %>%
  select("Year", "Month", "Day", "UTC_Hour", "PM2.5") %>%
  filter(Year >= 2014)

#Take daily average from the hourly 

Vc_avgpm <- Vc_pm2.5 %>%
  group_by(Year, Month, Day) %>%
  rename(year = Year, month = Month, day = Day) %>%
  summarize(PM_mean = mean(PM2.5))

Vc_avgpm$city <- "vancouver"
```

## Importing Temperature Data

DC daily temperatures
```{r}
DC_all <-ghcnd_search('USW00013743')

DC_tdaily <- DC_all$tavg
DC_tdaily <- DC_tdaily %>%
  select(id, tavg, date) %>%
  separate(.,date, into = c("year","month", "day"), sep = "-") %>%
  filter(year >= 2014) %>%
  rename('avgtemp' = 'tavg')

DC_tdaily$month <- as.numeric(DC_tdaily$month)
DC_tdaily$day <- as.numeric(DC_tdaily$day)
DC_tdaily$year <- as.numeric(DC_tdaily$year)

#change from celsius degrees of tenth to celsius
DC_tdaily$avgtemp <- DC_tdaily$avgtemp/10

#replacing station ids with city name
DC_tdaily$id <- gsub('USW00013743', 'washingtondc',DC_tdaily$id)

#renaming id column to city name
DC_tdaily <- DC_tdaily %>% rename('city' = 'id')
```

Kyoto daily temperatures
```{r}
Ky_all <-ghcnd_search('JA000047759')

Ky_tdaily <- Ky_all$tavg
Ky_tdaily <- Ky_tdaily %>%
  select(id, tavg, date) %>%
  separate(.,date, into = c("year","month", "day"), sep = "-") %>%
  filter(year >= 2014) %>%
  rename('avgtemp' = 'tavg')

Ky_tdaily$month <- as.numeric(Ky_tdaily$month)
Ky_tdaily$day <- as.numeric(Ky_tdaily$day)
Ky_tdaily$year <- as.numeric(Ky_tdaily$year)

#change from celsius degrees of tenth to celsius
Ky_tdaily$avgtemp <- Ky_tdaily$avgtemp/10

#replacing station ids with city name
Ky_tdaily$id <- gsub('JA000047759', 'kyoto',Ky_tdaily$id)

#renaming id column to city name
Ky_tdaily <- Ky_tdaily %>% rename('city' = 'id')
```

Liestal daily temperatures
```{r}
Li_all <-ghcnd_search('SZ000001940')

Li_tdaily <- Li_all$tavg
Li_tdaily <- Li_tdaily %>%
  select(id, tavg, date) %>%
  separate(.,date, into = c("year","month", "day"), sep = "-") %>%
  filter(year >= 2014) %>%
  rename('avgtemp' = 'tavg')

Li_tdaily$month <- as.integer(Li_tdaily$month)
Li_tdaily$day <- as.integer(Li_tdaily$day)
Li_tdaily$year <- as.integer(Li_tdaily$year)

#change from celsius degrees of tenth to celsius
Li_tdaily$tavg <- Li_tdaily$avgtemp/10

#replacing station ids with city name
Li_tdaily$id <- gsub('SZ000001940', 'liestal',Li_tdaily$id)

#renaming id column to city name
Li_tdaily <- Li_tdaily %>% rename('city' = 'id')
```

Vancouver daily temperatures
```{r}
Vc_all <-ghcnd_search('CA001108395')

Vc_tdaily <- Vc_all$tavg
Vc_tdaily <- Vc_tdaily %>%
  select(id, tavg, date) %>%
  separate(.,date, into = c("year","month", "day"), sep = "-") %>%
  filter(year >= 2014) %>%
  rename('avgtemp' = 'tavg')

Vc_tdaily$month <- as.integer(Vc_tdaily$month)
Vc_tdaily$day <- as.integer(Vc_tdaily$day)
Vc_tdaily$year <- as.integer(Vc_tdaily$year)


#change from celsius degrees of tenth to celsius
Vc_tdaily$tavg <- Vc_tdaily$avgtemp/10

#replacing station ids with city name
Vc_tdaily$id <- gsub('CA001108395', 'vancouver',Vc_tdaily$id)

#renaming id column to city name
Vc_tdaily <- Vc_tdaily %>% rename('city' = 'id')
```

## Visualization
```{r}

DC_tempbloom <- left_join(DC_tdaily, DC_bloom)

DC_tempbloom %>%
  mutate(date = as.Date(str_c(year,month,day, sep = "-")),
         month = format(date, "%b")) %>%
  filter(year != 2022,
         month %in% month.abb[1:3]) %>%
  ggplot() + 
  aes(factor(bloom_doy), avgtemp) +
  geom_boxplot() +
  facet_grid(factor(month, levels = month.abb) ~ .)
```

```{r}
DC_tempbloom %>%
  mutate(date = as.Date(str_c(year,month,day, sep = "-")),
         month = format(date, "%b")) %>%
  filter(year != 2022,
         month %in% month.abb[1:12], !month %in% month.abb[4:11]) %>%
  ggplot() +
#  geom_boxplot(aes(factor(bloom_doy), avgtemp)) +
#  geom_point(aes(bloom_doy, avgtemp)) +
  geom_smooth(aes(bloom_doy, avgtemp), method = "lm") +
  facet_wrap(factor(month, levels = c("Dec", "Jan", "Feb", "Mar")) ~ ., scales = "free", nrow = 2) +
  coord_flip()
```

```{r}
DC_tempbloom %>%
  mutate(date = as.Date(str_c(year,month,day, sep = "-")),
         date_filter = as.Date(str_c(2021,month,day, sep = "-")),
         month = format(date, "%b"),
         date_filter = if_else(month %in% month.abb[10:12], date_filter - 365, date_filter)) %>%
  filter(year != 2022,
         date_filter > as.Date("2020-12-15"),
         date_filter < as.Date("2021-04-15")
  ) %>%
  group_by(year) %>%
  summarize(month_avgtemp = mean(avgtemp),
            bloom_doy = first(bloom_doy)) %>%
  ggplot() + 
  aes(month_avgtemp, bloom_doy) +
  geom_point() +
  geom_smooth(method = "lm")
```
```{r}
DC_pmbloom %>%
  mutate(date = as.Date(str_c(year,month,day, sep = "-")),
         month = format(date, "%b")) %>%
  filter(year != 2022,
         month %in% month.abb[1:12], !month %in% month.abb[4:11]) %>%
  ggplot() + 
  aes(bloom_doy, PM_mean) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(factor(month, levels = month.abb) ~ .)
```
```{r}
DC_pmbloom %>%
  mutate(date = as.Date(str_c(year,month,day, sep = "-")),
         date_filter = as.Date(str_c(2021,month,day, sep = "-")),
         month = format(date, "%b"),
         date_filter = if_else(month %in% month.abb[10:12], date_filter - 365, date_filter)) %>%
  filter(year != 2022,
          date_filter > as.Date("2020-12-15"),
         date_filter < as.Date("2021-04-15")
  ) %>%
  group_by(year) %>%
  summarize(bloom_doy = first(bloom_doy),
            month_PMavg = mean(PM_mean)) %>%
  ggplot() + 
  aes(month_PMavg, bloom_doy) +
  geom_point() +
  geom_smooth(method = "lm")
```
--------------------------------------------------------------------------------
##Predictions

#DC Data
```{r}
DC_pmbloom <- left_join(DC_avgpm, DC_bloom)
DC_data <- full_join(DC_pmbloom, DC_tdaily)

DC_data$date<-as.Date(with(DC_data,paste(year,month,day,sep="-")),"%Y-%m-%d")
DC_data$week_number <- as.numeric(strftime(DC_data$date, format = "%V"))
DC_data$week <- floor_date(DC_data$date, "week")

DC_data <- DC_data %>%
  group_by(week) %>%
  mutate(weektemp = mean(avgtemp)) 

DC_data = DC_data[,!(names(DC_data) %in% c("location"))]
```

# DC Model and Predictions
```{r}
pred_temp <- lm(DC_data$weektemp ~ cos(DC_data$week_number*2*pi/52) + sin(DC_data$week_number*2*pi/52))
summary(pred_temp)

newtemp <- data.frame(predict(pred_temp, newdata = data.frame(date.int = 'year')))

#predict weektemp
ggplot(DC_data) +
  aes(week_number, weektemp) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ sin(x * 2 * pi / 52) + cos(x * 2 * pi / 52))


temp_pred <-
lm(weektemp ~ year + sin(week_number * 2 * pi / 52) + cos(week_number * 2 * pi / 52),
   data = DC_data)

predict(temp_pred, newdata = data.frame(week_number = 1:52, year = 2023))

#predict pm
ggplot(DC_data) +
  aes(week_number, PM_mean) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ sin(x * 2 * pi / 52) + cos(x * 2 * pi / 52)) +
  facet_wrap(~year)


PM_pred <-
  lm(PM_mean ~ year + sin(week_number * 2 * pi / 52) + cos(week_number * 2 * pi / 52),
     data = DC_data)

predictions <- 
  cbind(expand.grid(week_number = 1:52, year = 2023:2032),
        weektemp =
    predict(temp_pred, newdata = expand.grid(week_number = 1:52, year = 2023:2032)),
    PM_mean =
      predict(PM_pred, newdata = expand.grid(week_number = 1:52, year = 2023:2032)))

model <- lm(bloom_doy ~ PM_mean + weektemp + week_number + year, data = DC_data)

predictions$bloom_doy_pred <- predict(model, newdata = predictions)

DC_pred <- predictions %>%
  group_by(year) %>%
  summarize(bloom_doy = mean(bloom_doy_pred)) %>%
  mutate(bloom_doy = round(bloom_doy))

DC_pred
```

#Kyoto Data
```{r}
Ky_pmbloom <- left_join(Ky_avgpm, Ky_bloom)
Ky_data <- full_join(Ky_pmbloom, Ky_tdaily)

Ky_data$date<-as.Date(with(Ky_data,paste(year,month,day,sep="-")),"%Y-%m-%d")
Ky_data$week_number <- as.numeric(strftime(Ky_data$date, format = "%V"))
Ky_data$week <- floor_date(Ky_data$date, "week")

Ky_data <- Ky_data %>%
  group_by(week) %>%
  mutate(weektemp = mean(avgtemp)) 

Ky_data = Ky_data[,!(names(Ky_data) %in% c("location"))]
```

#Kyoto Model and Predictions
```{r}
#predict weektemp
ggplot(Ky_data) +
  aes(week_number, weektemp) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ sin(x * 2 * pi / 52) + cos(x * 2 * pi / 52))


temp_pred1 <-
lm(weektemp ~ year + sin(week_number * 2 * pi / 52) + cos(week_number * 2 * pi / 52),
   data = Ky_data)

predict(temp_pred1, newdata = data.frame(week_number = 1:52, year = 2023))

#predict pm
ggplot(Ky_data) +
  aes(week_number, PM_mean) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ sin(x * 2 * pi / 52) + cos(x * 2 * pi / 52)) +
  facet_wrap(~year)


PM_pred1 <-
  lm(PM_mean ~ year + sin(week_number * 2 * pi / 52) + cos(week_number * 2 * pi / 52),
     data = Ky_data)

predictions <- 
  cbind(expand.grid(week_number = 1:52, year = 2023:2032),
        weektemp =
    predict(temp_pred, newdata = expand.grid(week_number = 1:52, year = 2023:2032)),
    PM_mean =
      predict(PM_pred, newdata = expand.grid(week_number = 1:52, year = 2023:2032)))

model <- lm(bloom_doy ~ PM_mean + weektemp + week_number + year, data = Ky_data)

predictions$bloom_doy_pred <- predict(model, newdata = predictions)

Ky_pred <- predictions %>%
  group_by(year) %>%
  summarize(bloom_doy = mean(bloom_doy_pred)) %>%
  mutate(bloom_doy = round(bloom_doy))
  
Ky_pred
```

#Liestal Data
```{r}
Li_pmbloom <- left_join(Li_avgpm, Li_bloom)
Li_data <- full_join(Li_pmbloom, Li_tdaily)

Li_data$date<-as.Date(with(Li_data,paste(year,month,day,sep="-")),"%Y-%m-%d")
Li_data$week_number <- as.numeric(strftime(Li_data$date, format = "%V"))
Li_data$week <- floor_date(Li_data$date, "week")

Li_data <- Li_data %>%
  group_by(week) %>%
  mutate(weektemp = mean(avgtemp)) 

Li_data = Li_data[,!(names(Li_data) %in% c("location"))]
```

# Liestal Model and Predictions
```{r}
pred_temp <- lm(Li_data$weektemp ~ cos(Li_data$week_number*2*pi/52) + sin(Li_data$week_number*2*pi/52))
summary(pred_temp)

newtemp <- data.frame(predict(pred_temp, newdata = data.frame(date.int = 'year')))

#predict weektemp
ggplot(Li_data) +
  aes(week_number, weektemp) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ sin(x * 2 * pi / 52) + cos(x * 2 * pi / 52))


temp_pred <-
lm(weektemp ~ year + sin(week_number * 2 * pi / 52) + cos(week_number * 2 * pi / 52),
   data = Li_data)

predict(temp_pred, newdata = data.frame(week_number = 1:52, year = 2023))

#predict pm
ggplot(Li_data) +
  aes(week_number, PM_mean) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ sin(x * 2 * pi / 52) + cos(x * 2 * pi / 52)) +
  facet_wrap(~year)


PM_pred <-
  lm(PM_mean ~ year + sin(week_number * 2 * pi / 52) + cos(week_number * 2 * pi / 52),
     data = Li_data)

predictions <- 
  cbind(expand.grid(week_number = 1:52, year = 2023:2032),
        weektemp =
    predict(temp_pred, newdata = expand.grid(week_number = 1:52, year = 2023:2032)),
    PM_mean =
      predict(PM_pred, newdata = expand.grid(week_number = 1:52, year = 2023:2032)))

model <- lm(bloom_doy ~ PM_mean + weektemp + week_number + year, data = Li_data)

predictions$bloom_doy_pred <- predict(model, newdata = predictions)

Li_pred <- predictions %>%
  group_by(year) %>%
  summarize(bloom_doy = mean(bloom_doy_pred)) %>%
  mutate(bloom_doy = round(bloom_doy))

Li_pred
```

#Vancouver Data
```{r}
Vc_data <- full_join(Vc_avgpm, Vc_tdaily)

Vc_data$date<-as.Date(with(Vc_data,paste(year,month,day,sep="-")),"%Y-%m-%d")
Vc_data$week_number <- as.numeric(strftime(Vc_data$date, format = "%V"))
Vc_data$week <- floor_date(Vc_data$date, "week")

Vc_data <- Vc_data %>%
  group_by(week) %>%
  mutate(weektemp = mean(avgtemp)) 

Vc_data = Vc_data[,!(names(Vc_data) %in% c("location"))]
```

#Vancouver Model and Predictions
```{r}
pred_temp <- lm(Vc_data$weektemp ~ cos(Vc_data$week_number*2*pi/52) + sin(Vc_data$week_number*2*pi/52))
summary(pred_temp)

newtemp <- data.frame(predict(pred_temp, newdata = data.frame(date.int = 'year')))

#predict weektemp
ggplot(Vc_data) +
  aes(week_number, weektemp) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ sin(x * 2 * pi / 52) + cos(x * 2 * pi / 52))


temp_pred <-
lm(weektemp ~ year + sin(week_number * 2 * pi / 52) + cos(week_number * 2 * pi / 52),
   data = Vc_data)

predict(temp_pred, newdata = data.frame(week_number = 1:52, year = 2023))

#predict pm
ggplot(Vc_data) +
  aes(week_number, PM_mean) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ sin(x * 2 * pi / 52) + cos(x * 2 * pi / 52)) +
  facet_wrap(~year)


PM_pred <-
  lm(PM_mean ~ year + sin(week_number * 2 * pi / 52) + cos(week_number * 2 * pi / 52),
     data = Vc_data)

predictions <- 
  cbind(expand.grid(week_number = 1:52, year = 2023:2032),
        weektemp =
    predict(temp_pred, newdata = expand.grid(week_number = 1:52, year = 2023:2032)),
    PM_mean =
      predict(PM_pred, newdata = expand.grid(week_number = 1:52, year = 2023:2032)))

model <- lm(bloom_doy ~ PM_mean + weektemp + week_number + year, data = DC_data)

predictions$bloom_doy_pred <- predict(model, newdata = predictions)

Vc_pred <- predictions %>%
  group_by(year) %>%
  summarize(bloom_doy = mean(bloom_doy_pred)) %>%
  mutate(bloom_doy = round(bloom_doy))

Vc_pred
```

## Combine Predictions
```{r}

DC_pred <-DC_pred %>% rename('washingtondc' = 'bloom_doy')
Ky_pred <-Ky_pred %>% rename('kyoto' = 'bloom_doy')
Li_pred <-Li_pred %>% rename('liestal' = 'bloom_doy')
Vc_pred <-Vc_pred %>% rename('vancouver' = 'bloom_doy')

predictions <- full_join(Ky_pred, Li_pred)
predictions <- full_join(predictions, DC_pred)
predictions <- full_join(predictions, Vc_pred)

```

Export
```{r}
write.csv(predictions,"predictions2023.csv", row.names = FALSE)
```